---
title: "Demographic Feature Analysis for Diabetes Risk"
author: "Bereket Tarekegn"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

# Project Overview

This section of the ADS 503 final project focuses on demographic and socioeconomic predictors of diabetes risk using the CDC BRFSS 2015 Diabetes Health Indicators dataset.

The original outcome variable, `Diabetes_012`, has three classes: no diabetes, prediabetes, and diabetes. For this analysis, I converted the outcome into a binary screening target:

- `NoRisk` = no diabetes
- `AtRisk` = prediabetes or diabetes

This approach fits the project goal because the purpose is early screening. The model is not intended to diagnose diabetes. Instead, it is designed to identify people who may need follow-up screening or preventive care.

# Load Packages

```{r setup}
# Install packages first if needed:
# install.packages(c("tidyverse", "caret", "knitr", "rpart", "randomForest", "pROC"))

library(tidyverse)
library(caret)
library(knitr)
library(rpart)
library(randomForest)
library(pROC)

set.seed(503)
```

# Load Dataset

```{r load-data}
csv_path <- "C:/Project/diabetes_012_health_indicators_BRFSS2015.csv"

if (!file.exists(csv_path)) {
  csv_path <- "diabetes_012_health_indicators_BRFSS2015.csv"
}

if (!file.exists(csv_path)) {
  stop("Dataset not found. Please put diabetes_012_health_indicators_BRFSS2015.csv in C:/Project or in the same folder as this qmd file.")
}

diabetes <- read.csv(csv_path)

dim(diabetes)
names(diabetes)
```

The dataset contains 253,680 observations and 22 variables from the CDC BRFSS 2015 survey.

# Create Binary Diabetes Risk Target

```{r create-binary-target}
demo_data <- diabetes %>%
  mutate(
    Diabetes_Risk = ifelse(Diabetes_012 == 0, "NoRisk", "AtRisk"),
    Diabetes_Risk = factor(Diabetes_Risk, levels = c("AtRisk", "NoRisk"))
  ) %>%
  dplyr::select(
    Diabetes_Risk,
    Age,
    Sex,
    Education,
    Income,
    AnyHealthcare,
    NoDocbcCost
  )

str(demo_data)
summary(demo_data)
```

The target variable is now binary. Prediabetes and diabetes are combined into the `AtRisk` class.

# Recode Predictors

```{r recode-predictors}
demo_data <- demo_data %>%
  mutate(
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    Age = factor(Age),
    Education = factor(Education),
    Income = factor(Income),
    AnyHealthcare = factor(AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes")),
    NoDocbcCost = factor(NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes"))
  )

str(demo_data)
summary(demo_data)
```

These variables are treated as categorical features because they represent survey response groups.

# Missing Value Analysis

```{r missing-values}
missing_table <- data.frame(
  Variable = names(demo_data),
  Missing_Count = colSums(is.na(demo_data)),
  Missing_Percent = round(colSums(is.na(demo_data)) / nrow(demo_data) * 100, 2)
)

kable(missing_table, caption = "Missing Values for Demographic Feature Group")
```

The selected demographic variables do not contain missing values, so no imputation is needed for this feature group.

# Class Distribution

```{r class-distribution}
target_table <- demo_data %>%
  count(Diabetes_Risk) %>%
  mutate(Percent = round(n / sum(n) * 100, 2))

kable(target_table, caption = "Binary Diabetes Risk Class Distribution")
```

The class distribution is imbalanced. There are more respondents in the `NoRisk` group than in the `AtRisk` group. Because of this imbalance, accuracy alone is not enough to judge model performance.

# Exploratory Data Analysis

## Diabetes Risk Distribution

```{r diabetes-risk-plot}
ggplot(demo_data, aes(x = Diabetes_Risk, fill = Diabetes_Risk)) +
  geom_bar() +
  labs(
    title = "Distribution of Diabetes Risk Status",
    x = "Diabetes Risk Status",
    y = "Number of Respondents",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Age

```{r age-risk-plot}
ggplot(demo_data, aes(x = Age, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Sex

```{r sex-risk-plot}
ggplot(demo_data, aes(x = Sex, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Sex",
    x = "Sex",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Education

```{r education-risk-plot}
ggplot(demo_data, aes(x = Education, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Education Level",
    x = "Education Level",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Income

```{r income-risk-plot}
ggplot(demo_data, aes(x = Income, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Income Group",
    x = "Income Group",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Healthcare Access

```{r healthcare-risk-plot}
ggplot(demo_data, aes(x = AnyHealthcare, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Healthcare Access",
    x = "Has Healthcare Coverage",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

## Diabetes Risk by Medical Cost Barrier

```{r cost-risk-plot}
ggplot(demo_data, aes(x = NoDocbcCost, fill = Diabetes_Risk)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Risk by Medical Cost Barrier",
    x = "Could Not See Doctor Because of Cost",
    y = "Proportion",
    fill = "Risk Status"
  ) +
  theme_minimal()
```

# Train and Test Split

The dataset has more than 250,000 observations, so a train and test split is appropriate. The split is stratified to preserve the class distribution.

```{r train-test-split}
set.seed(503)

train_index <- createDataPartition(
  demo_data$Diabetes_Risk,
  p = 0.80,
  list = FALSE
)

train_data <- demo_data[train_index, ]
test_data <- demo_data[-train_index, ]

prop.table(table(train_data$Diabetes_Risk)) * 100
prop.table(table(test_data$Diabetes_Risk)) * 100
```

# Balanced Training Sample

The training data is imbalanced, so a balanced training sample is used to help the models learn both classes. The final evaluation still uses the original test set.

```{r balanced-training-sample}
set.seed(503)

at_risk_train <- train_data %>%
  filter(Diabetes_Risk == "AtRisk")

no_risk_train <- train_data %>%
  filter(Diabetes_Risk == "NoRisk") %>%
  slice_sample(n = nrow(at_risk_train))

train_balanced <- bind_rows(at_risk_train, no_risk_train) %>%
  slice_sample(prop = 1)

table(train_balanced$Diabetes_Risk)
```

# Cross Validation Setup

```{r cross-validation-setup}
set.seed(503)

ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 2,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)
```

The models are tuned using sensitivity because this project is framed as a health screening problem. For screening, catching true at-risk cases is important.

# Model 1: Logistic Regression

```{r logistic-model, cache=TRUE}
set.seed(503)

model_logit <- train(
  Diabetes_Risk ~ .,
  data = train_balanced,
  method = "glm",
  family = "binomial",
  trControl = ctrl,
  metric = "Sens"
)

model_logit
```

Logistic regression is used as a baseline model because it is simple and interpretable.

# Model 2: Decision Tree

```{r decision-tree-model, cache=TRUE}
set.seed(503)

model_tree <- train(
  Diabetes_Risk ~ .,
  data = train_balanced,
  method = "rpart",
  trControl = ctrl,
  metric = "Sens",
  tuneLength = 10
)

model_tree
plot(model_tree)
```

The decision tree can capture nonlinear relationships and is easy to explain visually.

# Model 3: Random Forest

```{r random-forest-model, cache=TRUE}
set.seed(503)

model_rf <- train(
  Diabetes_Risk ~ .,
  data = train_balanced,
  method = "rf",
  trControl = ctrl,
  metric = "Sens",
  tuneGrid = data.frame(mtry = c(2, 3, 4)),
  ntree = 100
)

model_rf
plot(model_rf)
```

Random Forest is included because it can capture more complex relationships and provides variable importance results.

# Cross Validation Comparison

```{r cross-validation-comparison}
model_results <- resamples(
  list(
    Logistic_Regression = model_logit,
    Decision_Tree = model_tree,
    Random_Forest = model_rf
  )
)

summary(model_results)
dotplot(model_results, metric = "Sens")
bwplot(model_results, metric = "Sens")
```

# Test Set Evaluation

```{r evaluation-function}
calculate_metrics <- function(predicted, actual, model_name) {
  cm <- confusionMatrix(predicted, actual, positive = "AtRisk")
  
  accuracy <- as.numeric(cm$overall["Accuracy"])
  kappa <- as.numeric(cm$overall["Kappa"])
  sensitivity <- as.numeric(cm$byClass["Sensitivity"])
  specificity <- as.numeric(cm$byClass["Specificity"])
  precision <- as.numeric(cm$byClass["Pos Pred Value"])
  f1 <- as.numeric(cm$byClass["F1"])
  balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)
  
  data.frame(
    Model = model_name,
    Accuracy = round(accuracy, 4),
    Kappa = round(kappa, 4),
    Sensitivity_Recall = round(sensitivity, 4),
    Specificity = round(specificity, 4),
    Precision = round(precision, 4),
    F1 = round(f1, 4),
    Balanced_Accuracy = round(balanced_accuracy, 4)
  )
}
```

```{r test-predictions}
pred_logit <- predict(model_logit, newdata = test_data)
pred_tree <- predict(model_tree, newdata = test_data)
pred_rf <- predict(model_rf, newdata = test_data)

cm_logit <- confusionMatrix(pred_logit, test_data$Diabetes_Risk, positive = "AtRisk")
cm_tree <- confusionMatrix(pred_tree, test_data$Diabetes_Risk, positive = "AtRisk")
cm_rf <- confusionMatrix(pred_rf, test_data$Diabetes_Risk, positive = "AtRisk")

cm_logit
cm_tree
cm_rf
```

# Model Comparison Table

```{r model-comparison-table}
comparison_table <- bind_rows(
  calculate_metrics(pred_logit, test_data$Diabetes_Risk, "Logistic Regression"),
  calculate_metrics(pred_tree, test_data$Diabetes_Risk, "Decision Tree"),
  calculate_metrics(pred_rf, test_data$Diabetes_Risk, "Random Forest")
)

kable(comparison_table, caption = "Test Set Model Comparison")
```

This table reports multiple metrics instead of only accuracy. This is important because the dataset is imbalanced.

# Variable Importance

```{r variable-importance}
importance_logit <- varImp(model_logit)
importance_tree <- varImp(model_tree)
importance_rf <- varImp(model_rf)

plot(importance_logit, main = "Variable Importance: Logistic Regression")
plot(importance_tree, main = "Variable Importance: Decision Tree")
plot(importance_rf, main = "Variable Importance: Random Forest")
```

Variable importance helps explain which demographic and socioeconomic variables are most useful for identifying diabetes risk.

# Discussion

This analysis used demographic and socioeconomic variables to predict diabetes risk. The target variable was converted from three classes into a binary screening target because the goal is to identify people who may be at risk and may need follow-up care.

The dataset is imbalanced, so accuracy alone is not enough to evaluate model quality. A model could appear accurate by predicting the majority class too often. For that reason, this analysis reports sensitivity, specificity, F1 score, balanced accuracy, and confusion matrices.

Sensitivity is important because this is a screening problem. If the model misses too many at-risk individuals, it would not be useful for early identification. Specificity is also important because too many false positives could create unnecessary follow-up work.

# Conclusion

The demographic feature group provides useful information for early diabetes risk screening. Variables such as age, sex, education, income, healthcare access, and medical cost barriers are practical because they can be collected without laboratory testing.

The results from this demographic analysis can be combined with the lifestyle feature analysis to support the final group project.
