---
title: "ADS 503 Final Project: Medical and Clinical Feature Analysis"
author: "Luisa Gonzalez"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This section of the group project focuses on medical and clinical health indicators from the CDC Diabetes Health Indicators dataset. The goal is to determine whether medical conditions and self-reported health measures can help predict diabetes status.

# Business Question

Can medical and clinical health indicators predict whether a person has no diabetes, prediabetes, or diabetes?
---

```{r setup}
# Load Packages
library(tidyverse)
library(caret)
library(knitr)
library(nnet)
library(rpart)
library(pROC)

set.seed(503)

```

```{r}
#load dataset 
diabetes <- read.csv("diabetes_012_health_indicators_BRFSS2015.csv")

dim(diabetes)
names(diabetes)
str(diabetes)

```

```{r}
#Select Medical and Clinical Features
medical_data <- diabetes %>%
  select(
    Diabetes_012,
    HighBP,
    HighChol,
    CholCheck,
    BMI,
    Stroke,
    HeartDiseaseorAttack,
    GenHlth,
    MentHlth,
    PhysHlth,
    DiffWalk
  ) %>%
  mutate(
    Diabetes_012 = factor(
      Diabetes_012,
      levels = c(0, 1, 2),
      labels = c("No_Diabetes", "Prediabetes", "Diabetes")
    ),
    HighBP = factor(HighBP),
    HighChol = factor(HighChol),
    CholCheck = factor(CholCheck),
    Stroke = factor(Stroke),
    HeartDiseaseorAttack = factor(HeartDiseaseorAttack),
    DiffWalk = factor(DiffWalk),
    GenHlth = factor(GenHlth),
    BMI = as.numeric(BMI),
    MentHlth = as.numeric(MentHlth),
    PhysHlth = as.numeric(PhysHlth)
  )

str(medical_data)
table(medical_data$Diabetes_012, useNA = "always")
```

```{r}
#Target Variable Distribution 
target_table <- medical_data %>%
  count(Diabetes_012) %>%
  mutate(
    Percent = round(n / sum(n) * 100, 2)
  )

kable(
  target_table,
  caption = "Distribution of Diabetes Status"
)

ggplot(medical_data,
       aes(x = Diabetes_012)) +
  geom_bar() +
  labs(
    title = "Distribution of Diabetes Status",
    x = "Diabetes Status",
    y = "Number of Respondents"
  ) +
  theme_minimal()
```

#The outcome distribution shows that most respondents are classified as having no diabetes. Prediabetes represents the smallest group, while diabetes cases account for a substantially smaller proportion of observations than the no diabetes category. This class imbalance should be considered when evaluating model performance because accuracy alone may not fully reflect predictive ability.

```{r}
#EDA
#BMI plot 
ggplot(medical_data,
       aes(x = Diabetes_012,
           y = BMI,
           fill = Diabetes_012)) +
  geom_boxplot() +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI"
  ) +
  theme_minimal()

#Diabetes Status by General Health
ggplot(medical_data,
       aes(x = GenHlth,
           fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by General Health",
    x = "General Health Rating",
    y = "Proportion",
    fill = "Diabetes Status"
  ) +
  theme_minimal()

# Diabetes Status by High Blood Pressure
ggplot(medical_data,
       aes(x = HighBP,
           fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by High Blood Pressure",
    x = "High Blood Pressure",
    y = "Proportion",
    fill = "Diabetes Status"
  ) +
  theme_minimal()
```

```{r}
#Train/Test 
set.seed(503)

train_index <- createDataPartition(
  medical_data$Diabetes_012,
  p = 0.70,
  list = FALSE
)

train_data <- medical_data[train_index, ]
test_data <- medical_data[-train_index, ]

dim(train_data)
dim(test_data)

prop.table(table(train_data$Diabetes_012))
prop.table(table(test_data$Diabetes_012))
```

```{r}
#Model Training Setup
ctrl <- trainControl(
  method = "cv",
  number = 5,
  savePredictions = "final"
)


```

```{r}
#Model 1: Multinomial Logistic Regression 
set.seed(503)

model_multinom <- train(
  Diabetes_012 ~ .,
  data = train_data,
  method = "multinom",
  trControl = ctrl,
  trace = FALSE
)

model_multinom
```

```{r}
# Model 2: Decision Tree
set.seed(503)

model_tree <- train(
  Diabetes_012 ~ .,
  data = train_data,
  method = "rpart",
  trControl = ctrl,
  tuneLength = 10
)

model_tree
plot(model_tree)
```

```{r}
# Model 3: Random Forest Model

set.seed(503)

rf_sample_index <- createDataPartition(
  train_data$Diabetes_012,
  p = 0.25,
  list = FALSE
)

train_data_rf <- train_data[rf_sample_index, ]

model_rf <- train(
  Diabetes_012 ~ .,
  data = train_data_rf,
  method = "rf",
  trControl = ctrl,
  tuneGrid = data.frame(mtry = c(2, 4, 6)),
  ntree = 100
)

model_rf
```

```{r}
#Evaluate Multinom Model on Test Set 
pred_multinom <- predict(model_multinom, test_data)

confusionMatrix(
  pred_multinom,
  test_data$Diabetes_012
)

pred_tree <- predict(model_tree, test_data)

confusionMatrix(
  pred_tree,
  test_data$Diabetes_012
)

pred_rf <- predict(model_rf, test_data)

confusionMatrix(
  pred_rf,
  test_data$Diabetes_012
)


```

```{r}
model_results <- data.frame(
  Model = c(
    "Multinomial Logistic Regression",
    "Decision Tree",
    "Random Forest"
  ),
  Accuracy = c(
    0.8473,
    0.8474,
    0.8473
  ),
  Kappa = c(
    0.1776,
    0.1724,
    0.1634
  )
)

kable(
  model_results,
  caption = "Test Set Model Comparison"
)
```

```{r}
varImp(model_multinom)
```
